The benefits of reporting critical effect size values

### 2024-12-16 paper/paper.tex
final

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{critical effect size values, statistical significance, hypothesis testing, effect sizes\newline\indent Word count: 3912}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\typesectitle}}

@@

\abstract{%
Critical effect size values represent the smallest detectable effect that can reach statistical significance given a specific sample size and alpha level. It can be useful to calculate the critical effect size when designing a study, and evaluate whether such effects are plausible. Reporting critical effect size values may be useful when the sample size has not been planned a priori, or when there is uncertainty about the expected sample size that can be collected, when researchers plan to analyze the data with a statistical hypothesis test. To assist researchers in calculating critical effect size values we have developed an R package that allows researchers to report critical effect size values for group comparisons, correlations, linear regressions, and meta-analyses. Reflecting on critical effect size values could benefit researchers during the planning phase of the study by helping them to understand the limitations of their research design. Critical effect size values are also useful when evaluating studies performed by other researchers when a-priori power analyses were not performed, especially when non-significant results were observed.
}

@@

The ``critical effect size'' refers to the smallest effect size that can be statistically significant, given the test performed, the sample size, and the alpha level (Lakens, 2022). Consider a study that tests a bivariate correlation with \(n = 20\) participants and uses \(\alpha < 0.05\) as the threshold for statistical significance in a two-sided test. The critical effect size values for this test are Pearson's \(r = -0.44\) and \(r = 0.44\), which means that the statistical test will yield a significant result for effects larger than \(r = -0.44\) or \(r = 0.44\), while observed effect sizes between these two values will lead to a statistically non-significant result. In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference use significance levels in hypothesis tests to make claims about the presence or absence of effects. Critical effect size values can serve as a useful complement for interpreting study results, communicate the informational value of a study design, and contextualize the interpretation of non-significant results, particularly when statistical significance is relied upon for inference. Reporting critical effect size values is especially important when the sample size was not (or could not) be predetermined on the basis of a smallest effect size of interest.

@@

First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to rejecting the null hypothesis are in line with realistic expectations. If the sample size is small and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, researchers will realize they are not able to collect sufficient data to perform an informative hypothesis test. An a-priori power analysis would typically lead to a similar conclusion, but reporting critical effect size values will focus the attention more strongly on which effect sizes are reasonable to expect. \textcolor{red}{In case of already conducted studies with small sample sizes it could be argued that it would be more informative to use retrospective design analysis, but this would require both knowledge on the plausible effect size and/or on the smallest effect size of interest (SESOI), which are not always easly to determine. The use of critical effect size values can be used in a simple an efficient way to evluate which findings could have not be found significant due to sample size limitations and based on the complexity of the test.}

@@

\textcolor{red}{To provide a more tangible illustration of the practical applications of the critical effect sizes, let us examine two real-world instances drawn from published research. The first study was conducted on musicians and non musicians to detect differences in working memory, a modest sample size of 57 participants (42 musicians and 15 controls) was collected (@weiss2014spatial). The two groups were compared on a variety of tasks, related to verbal working memory (VWM) and tasks related to auditory skills. Regarding the VWM, the syllable-span task was administered and scores were computed for the maximal span and for the total number of sequences they correctly repeated. A two-tailed $t$-test was conducted to compare the groups. The resulting effect was statistically significant for the maximal span, $t(55) = -2.5$, $p = .017$, the Cohen's d calculated on summary statistics is of $-.74$, critical $d = \pm .62$. But for the total number of sequences the test did not reach statistical significance ($t(55) = -1.8$, $p = 0.076$), actual $d = -0.31$ and critical $d$ again $\pm.62$. Thus, given the limited sample size, a result must be of rather large magnitude to be associated with $p < 0.05$, while medium to small effect sizes will simply fail to reach significance. This case makes it obvious that a failure to find a significant result does not imply absence of an effect, while a statistically significant effect may represent an overestimation. } Non-significant effects are likely, even if there is a true effect of interest. If a significant effect is observed, but most effects in a field are smaller than a study can detect, it is probable the effect size is overestimated (Hedges, 1984), and this should be taken into account when interpreting the effect size.
\textcolor{red}{Beyond critical effect size values, researchers should reason before conducting a study on what conclusions can be drawn, and for those cases where sample size constrains will make it impossible to find an effect because it is likely smaller than the critical effect size value, alternative options for data collection, such as multi-lab studies (\cite{AAA}), should be explored.}

@@

The second real-world instance involves the study by Kramer et al. (2014), which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of \(n = 689.003\), the researchers observed a statistically significant increase in the number of negative words typed by participants when positive posts were reduced in their timeline, \(t(310,044) = −5.63\), \(p < 0.001\) (non-directional test). Based on this and other results from the study, the authors conclude that \textcolor{red}{"the emotions expressed by friends, via online social networks, influence our own moods, constituting, to our knowledge, the first experimental evidence for massive-scale emotional contagion via social networks".} The critical values for Cohen's \(d\) in this experimental design are \(-0.006\) and \(+0.006\). It should have been clear from the outset that the researchers should have considered the effect size beyond merely testing for the statistical significance. The critical effect size should prompt reflection on which effect sizes are truly meaningful in this context, and whether the observed effect size (\(d = 0.02\)) is theoretically or practically interesting.

@@

@@

\color{blue} In this section, we provide guidance and formulas for computing standardized critical effect size with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's \(d\)), correlations (Pearson's \(r\)), and raw and standardized coefficients in linear models. The same equations can be used to calculated unstandardized effect sizes that can be preferred in some situations (e.g., linear models). \color{black} These formulas have been incorporated into R functions of the package \texttt{criticalESvalue} accessible at: \url{https://github.com/psicostat/criticalESvalue}, and are elaborated upon in the subsequent section.

@@

Where \(b\) is the unstandardized effect size that depends on the type of test. \color{blue} For example, in the two sample \(t\)-test the numerator is the difference between the two sample means while in the one-sample case is the difference between the sample mean and the population value. The denominator is the standard error of the numerator that depends on the sample size and the samples variances. \color{black}

\color{blue}

Equation \eqref{eq:eq-gen-d} describe a general formulation of a standardized effect size measure. Where \(b\) is the unstandardized effect size (e.g., the difference between two means) and \(s\) is the standardization term \color{black}. For example, in the two-samples case, \(s\) is the pooled standard deviation between the two samples, while in the paired samples case, \(s\) is the standard deviation of the differences for paired samples.

@@

Real-case scenarios may not always be that simple. Hence, we chose to expand the application of computing critical effect size values beyond Cohen's \(d\) and correlation to include linear regression with both raw and standardized coefficients. This serves as a first step in computing critical effect size values for a wider array of effects encountered in practical scenarios, where linear models and their extensions are commonly utilized for modeling purposes. A prerequisite is that researchers must be able to identify what parameters in their statistical models reflect the effect sizes of interest, and that they are able to assess their relevance. Notably, however, this prerequisite aligns with the requirements of APA style guidelines concerning the reporting of effect sizes. For further illustration and application, additional examples are provided in the Supplementary online material.

@@

@@

@@

@@

@@

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{critical effect size values, statistical significance, hypothesis testing, effect sizes\newline\indent Word count: 4676}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\typesectitle}}

@@

\abstract{%
Critical effect size values represent the smallest detectable effect that can reach statistical significance given a specific sample size, alpha level and test statistic. It can be useful to calculate the critical effect size when designing a study, and evaluate whether such effects are plausible. Reporting critical effect size values may be useful when the sample size has not been planned a priori, or when there is uncertainty about the expected sample size that can be collected, when researchers plan to analyze the data with a statistical hypothesis test. To assist researchers in calculating critical effect size values we have developed an R package that allows researchers to report critical effect size values for group comparisons, correlations, linear regressions, and meta-analyses. Reflecting on critical effect size values could benefit researchers during the planning phase of the study by helping them to understand the limitations of their research design. Critical effect size values are also useful when evaluating studies performed by other researchers when a-priori power analyses were not performed, especially when non-significant results are observed.
}

@@

The ``critical effect size'' refers to the smallest effect size that can be statistically significant, given the test performed, the sample size, and the alpha level (Lakens, 2022). Consider a study that tests a bivariate correlation with \(n = 20\) participants and uses \(\alpha < 0.05\) as the threshold for statistical significance in a two-sided test. The critical effect size values for this test are Pearson's \(r = -0.44\) and \(r = 0.44\), which means that the statistical test will yield a significant result for effects larger than \(r = -0.44\) or \(r = 0.44\), while observed effect sizes between these two values will lead to a statistically non-significant result. In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference using significance levels in hypothesis tests to make claims about the presence or absence of effects. Critical effect size values can serve as a useful complement for interpreting study results, communicate the informational value of a study design, and contextualize the interpretation of non-significant results, particularly when statistical significance is relied upon for inference. Reporting critical effect size values is especially important when the sample size was not (or could not) be predetermined on the basis of a smallest effect size of interest.

@@

First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to rejecting the null hypothesis are in line with realistic expectations. If the sample size is small and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, researchers will realize they are not able to collect sufficient data to perform an informative hypothesis test. An a-priori power analysis would typically lead to a similar conclusion, but reporting critical effect size values will focus the attention more strongly on which effect sizes are reasonable to expect. In case of already conducted studies with small sample sizes it could be argued that it would be more informative to use retrospective design analysis, but this would require both knowledge on the plausible effect size and/or on the smallest effect size of interest (SESOI), which are not always easly to determine. The use of critical effect size values can be used in a simple an efficient way to evluate which findings could have not be found significant due to sample size limitations and based on the complexity of the test. Power analysis provides information about how likely it is to detect a specific effect size if it truly exists in the population. However, it does not indicate the minimum effect size required to reach statistical significance, which could provide additional insights into the strengths and limitations of a study design.

@@

To provide a more tangible illustration of the practical applications of the critical effect sizes, let us examine a real-world instance drawn from published research. The study of interest was conducted on musicians and non musicians to detect differences in working memory, a modest sample size of 57 participants (42 musicians and 15 controls) was collected (Weiss et al. (2014)). The two groups were compared on a variety of tasks, related to verbal working memory (VWM) and tasks related to auditory skills. Regarding the VWM, the syllable-span task was administered and scores were computed for the maximal span and for the total number of sequences they correctly repeated. A two-tailed \(t\)-test was conducted to compare the groups. The resulting effect was statistically significant for the maximal span, \(t(55) = -2.5\), \(p = .017\), the Cohen's d calculated on summary statistics is of \(-.74\), critical \(d = \pm .62\). But for the total number of sequences the test did not reach statistical significance (\(t(55) = -1.8\), \(p = 0.076\)), actual \(d = -0.31\) and critical \(d\) again \(\pm.62\). Thus, given the limited sample size, a result must be of rather large magnitude to be associated with \(p < 0.05\), while medium to small effect sizes will simply fail to reach significance. This case makes it obvious that a failure to find a significant result does not imply absence of an effect, while a statistically significant effect may represent an overestimation. Non-significant effects in such scenarios are likely, even if there is a true effect of interest. If a significant effect is observed, but most effects in a field are smaller than a study can detect, it is probable the effect size is overestimated (Hedges, 1984), and this should be taken into account when interpreting the effect size.

Beyond critical effect size values, researchers should reason before conducting a study on what conclusions can be drawn with their experimental design, and for those cases where sample size constrains will make it impossible to find significance for a plausible effect because it is likely smaller than the critical effect size value, alternative options for data collection, such as multisite studies (Byers-Heinlein et al., 2020; Jarke et al., 2022; Moshontz et al., 2018; Sirois et al., 2023), should be explored.

@@

Consider now another real-world instance involving the study by Kramer et al. (2014), which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of \(n = 689.003\), the researchers observed a statistically significant increase in the number of negative words typed by participants when positive posts were reduced in their timeline, \(t(310,044) = −5.63\), \(p < 0.001\) (non-directional test). Based on this and other results from the study, the authors conclude that ``the emotions expressed by friends, via online social networks, influence our own moods, constituting, to our knowledge, the first experimental evidence for massive-scale emotional contagion via social networks'' (Kramer et al., 2014). The critical values for Cohen's \(d\) in this experimental design are \(-0.006\) and \(+0.006\). It should have been clear from the outset that the researchers should have considered the effect size beyond merely testing for the statistical significance. The critical effect size should prompt reflection on which effect sizes are truly meaningful in this context, and whether the observed effect size (\(d = 0.02\)) is theoretically or practically interesting.

@@

Lastly, routinely reasoning about critical effect sizes, alongside the use of commonly known practices to enhance the quality of research (i.e.~power analysis, design analysis, data simulation etc.), will bring benefits to researchers by making clear that significance is strictly related to the sample size, the alpha level and the statistic test used to analyse the data. Nevertheless, in educational settings, this concept will help students to better grasp such concepts and therefore give a more critical approach to published research and for their future studies.

@@

In this section, we provide guidance and formulas for computing standardized critical effect sizes with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's \(d\)), correlations (Pearson's \(r\)), and raw and standardized coefficients in linear models. The same equations can be used to calculated unstandardized effect sizes that can be preferred in some situations (e.g., linear models). These formulas have been incorporated into R functions of the package \texttt{criticalESvalue} accessible at: \url{https://github.com/psicostat/criticalESvalue}, and are elaborated upon in the subsequent section.

@@

Where \(b\) is the unstandardized effect size that depends on the type of test. For example, in the two sample \(t\)-test the numerator is the difference between the two sample means while in the one-sample case is the difference between the sample mean and the population value. The denominator is the standard error of the numerator that depends on the sample size and the samples variances.

Equation \eqref{eq:eq-gen-d} describes a general formulation of a standardized effect size measure. Where \(b\) is the unstandardized effect size (e.g., the difference between two means) and \(s\) is the standardization term. For example, in the two-samples case, \(s\) is the pooled standard deviation between the two samples, while in the paired samples case, \(s\) is the standard deviation of the differences for paired samples.

@@

Real-case scenarios may not always be that simple. Hence, we chose to expand the application of computing critical effect size values beyond Cohen's \(d\) and correlation to include linear regression with both raw and standardized coefficients and meta-analysis. This serves as a first step in computing critical effect size values for a wider array of effects encountered in practical scenarios, where linear models and their extensions are commonly utilized for modeling purposes. A prerequisite is that researchers must be able to identify what parameters in their statistical models reflect the effect sizes of interest, and that they are able to assess their relevance. Notably, however, this prerequisite aligns with the requirements of APA style guidelines concerning the reporting of effect sizes. For further illustration and application, additional examples are provided in the Supplementary online material.

@@

\leavevmode\vadjust pre{\hypertarget{ref-byers2020building}{}}%
Byers-Heinlein, K., Bergmann, C., Davies, C., Frank, M. C., Hamlin, J. K., Kline, M., Kominsky, J. F., Kosie, J. E., Lew-Williams, C., Liu, L., et al. (2020). Building a collaborative psychological science: Lessons learned from ManyBabies 1. \emph{Canadian Psychology/Psychologie Canadienne}, \emph{61}(4), 349.

@@

\leavevmode\vadjust pre{\hypertarget{ref-jarke2022roadmap}{}}%
Jarke, H., Anand-Vembar, S., Alzahawi, S., Andersen, T. L., Bojanić, L., Carstensen, A., Feldman, G., Garcia-Garzon, E., Kapoor, H., Lewis, S., et al. (2022). A roadmap to large-scale multi-country replications in psychology. \emph{Collabra: Psychology}, \emph{8}(1), 57538.

@@

\leavevmode\vadjust pre{\hypertarget{ref-moshontz2018psychological}{}}%
Moshontz, H., Campbell, L., Ebersole, C. R., IJzerman, H., Urry, H. L., Forscher, P. S., Grahe, J. E., McCarthy, R. J., Musser, E. D., Antfolk, J., et al. (2018). The psychological science accelerator: Advancing psychology through a distributed collaborative network. \emph{Advances in Methods and Practices in Psychological Science}, \emph{1}(4), 501--515.

@@

\leavevmode\vadjust pre{\hypertarget{ref-sirois2023pupil}{}}%
Sirois, S., Brisson, J., Blaser, E., Calignano, G., Donenfeld, J., Hepach, R., Hochmann, J.-R., Kaldy, Z., Liszkowski, U., Mayer, M., et al. (2023). The pupil collaboration: A multi-lab, multi-method analysis of goal attribution in infants. \emph{Infant Behavior and Development}, \emph{73}, 101890.

@@

\leavevmode\vadjust pre{\hypertarget{ref-weiss2014spatial}{}}%
Weiss, A. H., Biron, T., Lieder, I., Granot, R. Y., \& Ahissar, M. (2014). Spatial vision is superior in musicians when memory plays a role. \emph{Journal of Vision}, \emph{14}(9), 18--18.
